### 2025-09-12 YEAR2/MAT292_LectureNotes/main.tex
911

\subsection{Fixed Points}

@@

    To analyze the stability of the fixed points, we first find the fixed points by setting \( \frac{dx}{dt} = 0 \):
    $$\alpha x(1 - x) = 0 \implies x = 0 \text{ or } x = 1
    $$
    Next, we compute the derivative of \( f(x) = \alpha x(1 - x) \):
    $$\frac{df}{dx} = \alpha (1 - 2x)
    $$
    Evaluating this derivative at the fixed points, assumeing \( \alpha > 0 \):
    \begin{itemize}
        \item At \( x = 0 \):
        $$\frac{df}{dx} \bigg|_{x=0} = \alpha > 0 \quad \text{(unstable)}
        $$
        \item At \( x = 1 \):
        $$\frac{df}{dx} \bigg|_{x=1} = -\alpha < 0 \quad \text{(stable)}
        $$
    \end{itemize}
    Thus, the fixed point at \( x = 0 \) is unstable, while the fixed point at \( x = 1 \) is stable.
\end{example}
\subsubsection{Integrating Factor Method for First-Order Linear ODEs}
\begin{definition}[First-Order Linear ODE]
    A first-order linear ordinary differential equation (ODE) is an equation of the form:
    \begin{equation}
        \frac{dy}{dx} + P(x)y = Q(x)
    \end{equation}
    where \( P(x) \) and \( Q(x) \) are functions of the independent variable \( x \).
\end{definition}

@@

>>>>

\section{Qualitative Analysis of ODEs}
\subsection{Fixed Points}

@@

    To analyze the stability of the fixed points, we first find the fixed points by setting \( \frac{dx}{dt} = 0 \):
    $$\alpha x(1 - x) = 0 \implies x = 0 \text{ or } x = 1
    $$
    Next, we compute the derivative of \( f(x) = \alpha x(1 - x) \):
    $$\frac{df}{dx} = \alpha (1 - 2x)
    $$
    Evaluating this derivative at the fixed points, assumeing \( \alpha > 0 \):
    \begin{itemize}
        \item At \( x = 0 \):
        $$\frac{df}{dx} \bigg|_{x=0} = \alpha > 0 \quad \text{(unstable)}
        $$
        \item At \( x = 1 \):
        $$\frac{df}{dx} \bigg|_{x=1} = -\alpha < 0 \quad \text{(stable)}
        $$
    \end{itemize}
    Thus, the fixed point at \( x = 0 \) is unstable, while the fixed point at \( x = 1 \) is stable.
\end{example}
\subsubsection{First-Order Linear ODEs - Integrating Factor Method}
\begin{definition}[First-Order Linear ODE]
    A first-order linear ordinary differential equation (ODE) is an equation of the form:
    \begin{equation}
        \frac{dy}{dx} + P(x)y = Q(x)
    \end{equation}
    where \( P(x) \) and \( Q(x) \) are functions of the independent variable \( x \).
\end{definition}

@@

\subsection{Bifurcations}
\begin{definition}[Bifurcation]
    A bifurcation in a dynamical system occurs when a small change in the system's parameters causes a sudden qualitative change in its behavior.

    This can lead to the emergence or disappearance of fixed points, changes in stability, or the onset of periodic or chaotic behavior.
\end{definition}

\begin{definition}[Bifurcation Point]
    A bifurcation point is a specific value of a parameter in a dynamical system at which the qualitative behavior of the system changes. At this point, the system may undergo a bifurcation, leading to changes in the number or stability of fixed points, periodic orbits, or other dynamical features.

\end{definition}

\paragraph{Types of Bifurcations} There are several common types of bifurcations in dynamical systems, including:
    \begin{itemize}
        \item Saddle-Node Bifurcation: Two fixed points (one stable and one unstable) collide and annihilate each other as a parameter is varied.
        \item Transcritical Bifurcation: Two fixed points exchange their stability as a parameter is varied.
        \item Pitchfork Bifurcation: A single fixed point splits into three fixed points (one stable and two unstable, or vice versa) as a parameter is varied.
        \item Hopf Bifurcation: A fixed point loses stability and a small amplitude limit cycle (periodic orbit) emerges as a parameter is varied.
    \end{itemize}

\begin{example}
    Consider the ODE:
    $$
    x' = ax(1 - x)
    $$
    This is a classic example of a bifurcation scenario, where the behavior of the system changes qualitatively as the parameter \(a\) is varied. Consider the fix points of the system:
    $$
    x^* = 0, \quad x^* = 1
    $$
    So, we have:
    \begin{itemize}
        \item For \( a < 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = a < 0 \))
            \item \( x^* = 1 \) is unstable (since \( f'(1) = -a > 0 \))
        \end{itemize}
        \item For \( a > 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is unstable (since \( f'(0) = a > 0 \))
            \item \( x^* = 1 \) is stable (since \( f'(1) = -a < 0 \))
        \end{itemize}
        \item For \( a = 0 \):
        \begin{itemize}
            \item Both \( x^* = 0 \) and \( x^* = 1 \) are semi-stable (since \( f'(0) = 0 \) and \( f'(1) = 0 \))
        \end{itemize}
        This indicates a bifurcation at \( a = 0 \), where the stability of the fixed points changes.
    \end{itemize}
\end{example}

\begin{example}[Saddle-Node Bifurcation] \label{ex:bifurcation-saddle-node}
    Consider the ODE:
    $$
    x' = x^2 + a
    $$
    This equation has a family of fixed points given by:
    $$
    x^* = \pm\sqrt{-a}
    $$
    For \( a < 0 \), we have two real fixed points, while for \( a > 0 \), there are no real fixed points. At \( a = 0 \), the fixed points collide and disappear, indicating a bifurcation point at \( a = 0 \).
\end{example}

\begin{definition}[Bifurcation Diagram]
    A bifurcation diagram is a visual representation that illustrates how the fixed points of a dynamical system change as a parameter is varied.

    It is a plot of the fixed points against the parameter, showing regions of stability and instability, and we denote stable fixed points with solid lines and unstable fixed points with dashed lines. A bifurcation diagram of Example \ref{ex:bifurcation-saddle-node} is shown below:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                axis lines = middle,
                xlabel = \( a \),
                ylabel = \( x^* \),
                ymin = -2, ymax = 2,
                xmin = -2, xmax = 2,
                domain=-2:2,
                samples=100,
                width=10cm,
                height=6cm,
                grid=both,
                minor tick num=1,
                major grid style={line width=.2pt,draw=gray!50},
                minor grid style={line width=.1pt,draw=gray!20},
            ]
            % Stable fixed points
            \addplot[blue, thick] {-sqrt(-x)}; % Stable branch
            \addplot[red, dashed, thick] {sqrt(-x)}; % Stable branch

            % Bifurcation point
            \node at (axis cs:0,0) [circle,fill,inner sep=1.5pt]{};
            \node at (axis cs:0.3,0.3) {\( (0,0) \)};

            % Labels
            \node at (axis cs:-1,1) [anchor=south west] {unstable};
            \node at (axis cs:-1,-1) [anchor=north west] {stable};
            \end{axis}
        \end{tikzpicture}
        \caption{Bifurcation Diagram for \( x' = x^2 + a \)}
        \label{fig:bifurcation-diagram}
    \end{figure}

\end{definition}

\begin{example}[Pitchfork Bifurcation]
    Consider the ODE:
    $$
    x' = ax - x^3
    $$
    This equation has a family of fixed points given by:
    $$
    x^* = 0, \quad x^* = \pm\sqrt{a}
    $$
    For \( a \le 0 \), there is one real fixed point at \( x^* = 0 \). For \( a > 0 \), there are three real fixed points: \( x^* = 0 \) (unstable) and \( x^* = \pm\sqrt{a} \) (stable). At \( a = 0 \), the fixed point at \( x^* = 0 \) changes stability, indicating a bifurcation point at \( a = 0 \).

    We can plot the bifurcation diagram for this system as follows:

    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                axis lines = middle,
                xlabel = \( a \),
                ylabel = \( x^* \),
                ymin = -2, ymax = 2,
                xmin = -2, xmax = 2,
                domain=-2:2,
                samples=100,
                width=10cm,
                height=6cm,
                grid=both,
                minor tick num=1,
                major grid style={line width=.2pt,draw=gray!50},
                minor grid style={line width=.1pt,draw=gray!20},
            ]
            % Stable fixed point for a <= 0
            \addplot[blue, thick, domain=-2:0] {0}; % Stable branch for a <= 0
            % Unstable fixed point for a > 0
            \addplot[red, dashed, thick, domain=0:2] {0}; % Unstable branch for a > 0
            % Stable fixed points for a > 0
            \addplot[blue, thick] {sqrt(x)}; % Stable branch for a > 0
            \addplot[blue, thick] {-sqrt(x)}; % Stable
            % Bifurcation point
            \node at (axis cs:0,0) [circle,fill,inner sep=1.5pt]{};
            \node at (axis cs:0.3,0.3) {\( (0,0) \)};
            % Labels
            \node at (axis cs:-1,0.2) [anchor=south west] {stable};
            \node at (axis cs:0.9,1.1) [anchor=south west] {stable};
            \node at (axis cs:0.9,-1.1) [anchor=north west] {stable};
            \node at (axis cs:0.5,0.1) [anchor=south west] {unstable};
            \end{axis}
        \end{tikzpicture}
        \caption{Bifurcation Diagram for \( x' = ax - x^3 \)}
        \label{fig:pitchfork-bifurcation-diagram}
    \end{figure}
\end{example}

\begin{example}[Transcritical Bifurcation] \label{ex:bifurcation-transcritical}
    Consider the ODE:
    $$
    x' = x(r - x)
    $$
    This equation has a family of fixed points given by:
    $$
    x^* = 0, \quad x^* = r
    $$
    For $r < 0$, \( x^* = 0 \) is stable. For $r > 0$, \( x^* = 0 \) is unstable and \( x^* = r \) is stable. At \( r = 0 \), the stability of the fixed point at \( x^* = 0 \) changes, indicating a bifurcation point at \( r = 0 \). The bifurcation diagram is as follows:
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                axis lines = middle,
                xlabel = \( r \),
                ylabel = \( x^* \),
                ymin = -2, ymax = 2,
                xmin = -2, xmax = 2,
                domain=-2:2,
                samples=100,
                width=10cm,
                height=6cm,
                grid=both,
                minor tick num=1,
                major grid style={line width=.2pt,draw=gray!50},
                minor grid style={line width=.1pt,draw=gray!20},
            ]
            % Stable fixed point for r < 0
            \addplot[blue, thick, domain=-2:0] {0}; % Stable branch for r < 0
            % Unstable fixed point for r < 0
            \addplot[red, dashed, thick, domain=-2:0] {x}; % Unstable branch for r < 0
            % Unstable fixed point for r > 0
            \addplot[red, dashed, thick, domain=0:2] {0}; % Unstable branch for r > 0
            % Stable fixed point for r > 0
            \addplot[blue, thick, domain=0:2] {x}; % Stable branch for r > 0
            % Bifurcation point
            \node at (axis cs:0,0) [circle,fill,inner sep=1.5pt]{};
            \node at (axis cs:0.3,0.3) {\( (0,0) \)};
            % Labels
            \node at (axis cs:-1,0.2) [anchor=south west] {stable};
            \node at (axis cs:1,1) [anchor=south west] {stable};
            \node at (axis cs:0.5,0.1) [anchor=south west] {unstable};
            \end{axis}
        \end{tikzpicture}
        \caption{Bifurcation Diagram for \( x' = x(r - x) \)}
        \label{fig:bifurcation-diagram-2}
    \end{figure}
\end{example}

\paragraph{Linear Stability Analysis and Normal Forms} Consider the ODE in the above example \ref{ex:bifurcation-transcritical}:
We can verify the stability of the fixed points by performing linear stability analysis. Consider $x \sim x^*$ and $r \sim r^*$, such that $|x - x^*| = \epsilon$ and $|r - r^*| = \delta$ for small $\epsilon, \delta > 0$. Then,
\begin{align*}
    f(x, r) &= f(x^*, r^*) + (x-x^*) f_x(x^*, r^*) + (r-r^*) f_r(x^*, r^*) + \\
    &\quad \frac{(x-x^*)^2}{2} f_{xx}(x^*, r^*) + O(\epsilon^3, \delta^2, \epsilon\delta) \\
    &\approx a(x - x^*)^2 + b(r - r^*)
\end{align*}
which we refer to as the normal form of the system. Here, \( a = \frac{1}{2} f_{xx}(x^*, r^*) \) and \( b = f_r(x^*, r^*) \). And that $f_x(x^*, r^*) = 0$ is called a saddle node.

\begin{example}
    Consider the ODE:
    $$
    x' = rx + x^3 - x^5
    $$
    We can find the fixed points by setting \( x' = 0 \), and we have
    $$
    x^\star = 0 \quad \text{or} \quad x^\star = \pm\sqrt{\frac{1 \pm \sqrt{1 + 4r}}{2}}
    $$
    Now, we see the bifurcation points are at \( r = -\frac{1}{4} \) and \( r = 0 \). We have:
    \begin{itemize}
        \item For \( r < -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
        \end{itemize}
        \item For \( r = -\frac{1}{4} \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = \pm\frac{1}{2} \) are semi-stable (since \( f'(\pm\frac{1}{2}) = 0 \))
        \end{itemize}
        \item For \( -\frac{1}{4} < r < 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is stable (since \( f'(0) = r < 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is unstable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) > 0 \))
        \end{itemize}
        \item For \( r = 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is semi-stable (since \( f'(0) = 0 \))
            \item \( x^* = -1 \) is stable (since \( f'(-1) < 0 \))
            \item \( x^* = 1 \) is stable (since \( f'(1) < 0 \))
        \end{itemize}
        \item For \( r > 0 \):
        \begin{itemize}
            \item \( x^* = 0 \) is unstable (since \( f'(0) = r > 0 \))
            \item \( x^* = -\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(-\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
            \item \( x^* = \sqrt{\frac{1 + \sqrt{1 + 4r}}{2}} \) is stable (since \( f'(\sqrt{\frac{1 + \sqrt{1 + 4r}}{2}}) < 0 \))
        \end{itemize}
    \end{itemize}
    The bifurcation diagram is as follows:
    \begin{figure}[h!]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            axis lines = middle,
            xlabel = {$r$},
            ylabel = {$x^\star$},
            ymin = -2, ymax = 2,
            xmin = -1, xmax = 1,
            samples=200,
            domain=-0.25:1,
            legend style={at={(1.05,1)}, anchor=north west}
        ]
        % Stable and unstable branches from formula
        \addplot[blue, thick, domain=-0.25:1] {sqrt((1 + sqrt(1+4*x))/2)};
        \addplot[red, dashed, domain=-0.25:1] {sqrt((1 - sqrt(1+4*x))/2)};
        \addplot[blue, thick, domain=-0.25:1] {-sqrt((1 + sqrt(1+4*x))/2)};
        \addplot[red, dashed, domain=-0.25:1] {-sqrt((1 - sqrt(1+4*x))/2)};

        % vertical black dashed line at r = -1/4
        \draw[dashed] (axis cs:-0.25,-2) -- (axis cs:-0.25,2);
        \node at (axis cs:-0.25,2.2) {\( r = -\frac{1}{4} \)};  
        % The trivial equilibrium at x=0
        \addplot[blue, thick, domain=-1:-0.0001] {0}; % stable for r<0
        \addplot[red, dashed, domain=0:1] {0};        % unstable for r>0

        % Mark bifurcation points
        \node at (axis cs:-0.25,0.75) [circle,fill,inner sep=1.5pt]{};
        \node at (axis cs:-0.25,-0.75) [circle,fill,inner sep=1.5pt]{};
        \node at (axis cs:0,0) [circle,fill,inner sep=1.5pt]{};

        \legend{stable, unstable}
        \end{axis}
    \end{tikzpicture}
    \caption{Bifurcation diagram for $x' = rx + x^3 - x^5$. 
    Blue = stable equilibria, red dashed = unstable equilibria.}
\end{figure}
